Modelling BCG vaccination in the UK

### 2019-08-06 rmds/02-background.Rmd
map height

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active TB in their lifetime.[@Houben:2016jp] Of this number around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, the risk factors, the treatment, the global impact, and the impact in England and Wales are explored.

@@

### Known risk factors

@@

TB is a global disease with an estimated 10.4 million new cases in 2016,[@TheWorldHealthOrganization:2018va] of which 4.3 million were missed by health systems.[@WHO:2018]  Global incidence rates have decreased year-on-year since the early 2000s, with an average year on year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

@@

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Incidence rates in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201"}
map_tb_burden(verbose = FALSE)
```

@@

TB incidence in England and Wales has decreased dramatically from a century ago (Figure \@ref(fig:plot-incidence)^[or see http://www.seabbs.co.uk/shiny/TB_England_Wales]) for an interactive dashboard). However, in the past several decades, incidence rates first stabilised and have since increased since their lowest point in the 1990s. In 2000, there were 6044 notified TB cases in England, increasing to a maximum of 8280 notified TB cases in 2011. Since then, notifications have declined year on year.[@PHE2016a] Figure \@ref(fig:plot-incidence) includes the interventions discussed above (Table \@ref(tab:tab-interventions)) and indicates that the introduction of several antibiotics and BCG vaccination in the 1950s may have led to an extended decrease in incidence.

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year on year since 2010 (when it was 49%).[@PHE2017] However, in 2016, 23.3% (420/1,800) of non-UK born cases had traveled outside the UK, with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

@@

Strain typing or whole genome sequencing is used to establish case clustering. This can be used to rule out transmission between cases but does not necessarily confirm transmission. Approximately 60% of cases cluster with at least one other case, and whilst this varies year on year, the fluctuations appear to be small (approximately 1-2%).[@PHE2017] Therefore interpreting any trend in TB transmission from the current strain typing data is difficult. Between 2010 and 2016, the median cluster size was 3 cases (range 2-244). In these clusters,  74.4% (2,141/2,878) consisted of less than 5 cases and only 8.8% of clusters had more than 10 cases [@PHE2017]. UK born cases were more likely to cluster than non-UK born cases (71.1%, 4,200/5,910 vs. 56.1%, 10,166/18,121).[@PHE2017]

>>>>

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active TB in their lifetime.[@Houben:2016jp] Of this number around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, TB risk factors, TB treatment, global impact, and the impact TB in England and Wales are explored.

@@

### Risk factors

@@

TB is a global disease with an estimated 10.4 million new cases in 2016,[@TheWorldHealthOrganization:2018va] of which 4.3 million were missed by health systems.[@WHO:2018]  Global incidence rates have decreased year-on-year since the early 2000s, with an average year-on-year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

@@

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Incidence rates in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201", fig.height = 4}
map_tb_burden(verbose = FALSE)
```

@@

TB incidence in England and Wales has decreased dramatically from a century ago (Figure \@ref(fig:plot-incidence)^[or see http://www.seabbs.co.uk/shiny/TB_England_Wales]) for an interactive dashboard). However, in the past several decades, incidence rates first stabilised and have since increased since their lowest point in the 1990s. In 2000, there were 6044 notified TB cases in England, increasing to a maximum of 8280 notified TB cases in 2011. Since then, notifications have declined year-on-year.[@PHE2016a] Figure \@ref(fig:plot-incidence) includes the interventions discussed above (Table \@ref(tab:tab-interventions)) and indicates that the introduction of several antibiotics and BCG vaccination in the 1950s may have led to an extended decrease in incidence.

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year-on-year since 2010 (when it was 49%).[@PHE2017] However, in 2016, 23.3% (420/1,800) of non-UK born cases had traveled outside the UK, with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

@@

Strain typing or whole genome sequencing is used to establish case clustering. This can be used to rule out transmission between cases but does not necessarily confirm transmission. Approximately 60% of cases cluster with at least one other case, and whilst this varies year-on-year, the fluctuations appear to be small (approximately 1-2%).[@PHE2017] Therefore interpreting any trend in TB transmission from the current strain typing data is difficult. Between 2010 and 2016, the median cluster size was 3 cases (range 2-244). In these clusters,  74.4% (2,141/2,878) consisted of less than 5 cases and only 8.8% of clusters had more than 10 cases [@PHE2017]. UK born cases were more likely to cluster than non-UK born cases (71.1%, 4,200/5,910 vs. 56.1%, 10,166/18,121).[@PHE2017]
